---
title: "Hg_data_ag"
author: "Denver Link"
date: "2024-11-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Library
```{r}
library(tidyverse)
library(mwlaxeref)
library(mnsentinellakes)
```


# Data
```{r}
#mercury
mn_dnr <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "allfish_data_04042024_JLO.csv"))

glifwc <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "glifwc_hg.csv"))

#covariate
mglp <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Covariate Data", "MGLP_landuse_clarity_color_mercury_MI_MN_WI.csv"))

#mn zm data
mn_zm <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Covariate Data", "ZM_MN_NHD.csv"))
```

# Ag
This aggregation assumes all glifwc samples in MN were duplicates in the state data - I only take the Wisconsin samples from GLIFWC. 
```{r}
#get glifwc and mn data together
glifwc_merge <- glifwc %>% 
  #for the sake of ease I'm going to assume all mn samples are duplicates and throw them out
  filter(!(state %in% c("MN", "WI")))

mn_dnr_merge <- mn_dnr %>% 
  #removes samples with no hg, length, or dow data
  filter(!(is.na(LGTHIN) | is.na(HGPPM) | is.na(DOWID))) %>%   
  #only taking lake samples
  filter(TYPE %in% c("Lake", "LAKE")) %>% 
  #fixing dowid
  mutate(DOWID = fixlakeid(DOWID)) %>% 
  #removing great lake
  filter(DOWID != "16000100") %>% 
  #generating nhdid
  local_to_nhdhr(from_colname = "DOWID", states = "mn") %>% 
  #adding prefix to nhdid
  mutate(nhdhr.id = paste0("nhdhr_", nhdhr.id)) %>%
  #adding lagos
  local_to_lagos(from_colname = "DOWID", states = "mn") %>% 
  #selecting columns for rowbind
  rename(waterbody = WATERWAY,
         year = YEARCOLL,
         spp_code = SPEC,
         length_in = LGTHIN,
         wt_lbs = WTLB,
         age = AGE,
         hg_ppm = HGPPM,
         state_id = DOWID,
         sample_type = ANATHG,
         num_fish_in_sample = NOFISH,
         sex = SEX,
         hg_code = HGCODE,
         hg_lab = HGLAB,
         hg_notes = HGCMMT,
         lagoslakeid = lagos.id) %>% 
  mutate(state = "MN",
         county = NA,
         sample_date = mdy(DATECOL2)) %>% 
  select(state,
         county,
         waterbody,
         state_id,
         lagoslakeid,
         nhdhr.id,
         lagoslakeid,
         sample_date,
         year,
         spp_code,
         sample_type,
         num_fish_in_sample,
         length_in,
         wt_lbs,
         age,
         sex,
         hg_ppm,
         hg_code,
         hg_lab,
         hg_notes)

hg_data <- bind_rows(glifwc_merge, mn_dnr_merge)
rm(mn_dnr,
   mn_dnr_merge,
   glifwc,
   glifwc_merge)

wi_hg <- read_csv(file.path("G:", 
                             "Shared drives", 
                             "Hansen Lab", 
                             "RESEARCH PROJECTS", 
                             "Statewide mercury modeling", 
                             "Data", 
                             "Mercury Data", 
                             "WI",
                             "wi_hg_26NOV2024.csv"))
wi_hg_merge <- wi_hg %>% 
  mutate(state = "WI", 
         waterbody = primary_site,
         state_id = gsub("_", "",wbic),
         state_id = case_when(state_id == "" ~ NA,
                              TRUE ~ state_id),
         state_id = as.numeric(state_id),
         year = year(sample_date),
         spp_code = sample_type,
         sample_type = sample_form,
         num_fish_in_sample = number_of_fish,
         length_in = avg_length_in,
         wt_lbs = avg_body_wt_kg*2.2,
         sex = gender,
         hg_ppm = mercury_ug_g,
         hg_code = NA,
         hg_lab = NA,
         hg_notes = comments) %>% 
  local_to_lagos(from_colname = "state_id", states = "wi") %>%
  local_to_nhdhr(from_colname = "state_id", states = "wi") %>% 
  rename(lagoslakeid = lagos.id) %>% 
  select(state,
         county,
         waterbody,
         state_id,
         lagoslakeid,
         nhdhr.id,
         sample_date,
         year,
         spp_code,
         sample_type,
         num_fish_in_sample,
         length_in,
         wt_lbs,
         age,
         sex,
         hg_ppm,
         hg_code,
         hg_lab,
         hg_notes)

hg_data <- bind_rows(wi_hg_merge, hg_data) 
rm(wi_hg, wi_hg_merge)
rm(mglp)

##########STOPPED RIGHT HERE, I have to figure out this join################


# read in covariate data from lagos
lagos <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Covariate Data", "lagos_covarites_all_lakes_mn_wi_mi_23oct2025.csv"))

lagos <- lagos %>% 
  filter(!is.na(lake_nhdid)) %>% 
  distinct(lake_nhdid, .keep_all = T)

hg_data.check <- hg_data %>% 
  filter(!is.na(nhdhr.id)) %>% 
  select(-state,
         -lagoslakeid) %>% 
  rename(lake_nhdid = nhdhr.id) %>% 
  left_join(lagos, by = c("lake_nhdid"))

#checking hg data
hg_data.check %>% 
  mutate(no_lagos = case_when(is.na(nlcd_openwater11_pct) ~ "y",
                              TRUE ~ "n")) %>% 
  filter(no_lagos == "y") %>% 
  distinct(lake_nhdid, lake_id, state)

# mglp_merge <- mglp %>% 
#   lagos_to_nhdhr(from_colname = "lagoslakeid") %>% 
#   mutate(nhdhr.id = paste0("nhdhr_", nhdhr.id)) %>% 
#   select(nhdhr.id,
#          lake_area_ha,
#          nws_area_ha,
#          clarity_m_Corson_Dosch,
#          secchi_depth_m_LAGOS_LIMNO,
#          colora_pcu_LAGOS_LIMNO,
#          colort_pcu_LAGOS_LIMNO,
#          nws_nlcd_devhi24_pct,
#          nws_nlcd_fordec41_pct,
#          nws_nlcd_wetemerg95_pct)
# 
# hg_data <- hg_data %>% 
#   left_join(mglp_merge, by = "nhdhr.id")
# rm(mglp,
#    mglp_merge)
# 
# #adding zm data
# mn_zm_merge <- mn_zm %>% 
#   filter(!is.na(MNDOW_ID)) %>% 
#   mutate(DOW = as.character(gsub("mndow_", "", MNDOW_ID)),
#          nhdid = NHD_ID) %>% 
#   select(DOW,
#          Year_Infested)  %>% 
#   mutate(zm_lag_year_infested = Year_Infested +2) %>% 
#   rename(state_id = DOW,
#          year_infested_zm = Year_Infested,
#          year_infested_zm_lag = zm_lag_year_infested) 
# 
# hg_data <- hg_data %>% 
#   left_join(mn_zm_merge, by = "state_id") %>% 
#   mutate(zm_lake = case_when(is.na(year_infested_zm) ~ "n",
#                              TRUE ~ "y"),
#          zm_sample_lag = case_when(year_infested_zm_lag <= year ~ "y",
#                                TRUE ~ "n"),
#          time_since_invasion = year - year_infested_zm)
# rm(mn_zm, mn_zm_merge)
```

#products
```{r}
#mn data with mglp and zm
mn_data <- hg_data %>% 
  filter(!is.na(state_id)) %>% 
  select(state, 
         county, 
         waterbody,
         state_id,
         nhdhr.id,
         lagoslakeid,
         lake_area_ha,
         nws_area_ha,
         clarity_m_Corson_Dosch,
         secchi_depth_m_LAGOS_LIMNO,
         colora_pcu_LAGOS_LIMNO,
         colort_pcu_LAGOS_LIMNO,
         nws_nlcd_devhi24_pct,
         nws_nlcd_fordec41_pct,
         nws_nlcd_wetemerg95_pct,
         zm_lake,
         year_infested_zm,
         year_infested_zm_lag,
         zm_sample_lag,
         time_since_invasion,
         sample_date,
         year,
         sample_type,
         num_fish_in_sample,
         spp_code,
         length_in,
         wt_lbs,
         sex,
         age,
         hg_ppm,
         hg_code,
         hg_lab,
         hg_notes)
write_csv(mn_data, "mn_lakes_hg_covariates.csv")
```

# data coverage
```{r}
hg_data %>% 
  #currently removing all na nhdhr so that I do not get glifwc data - only looking at mglp coverage in MN
  filter(!is.na(nhdhr.id)) %>% 
  distinct(nhdhr.id, 
           lake_area_ha,
           nws_area_ha,
           clarity_m_Corson_Dosch,
           secchi_depth_m_LAGOS_LIMNO,
           colora_pcu_LAGOS_LIMNO,
           colort_pcu_LAGOS_LIMNO,
           nws_nlcd_devhi24_pct,
           nws_nlcd_fordec41_pct,
           nws_nlcd_wetemerg95_pct) %>% 
  summarize(across(everything(), ~ mean(is.na(.)) * 100, .names = "na_count_{.col}")) %>% 
  pivot_longer(cols = everything(), 
               names_to = "parameter", 
               values_to = "na_percentage") %>%
  mutate(parameter = str_remove(parameter, "na_percent_"))

hg_data %>% 
  filter(!is.na(nhdhr.id)) %>% 
  filter(is.na(nws_nlcd_fordec41_pct)) %>% 
  distinct(waterbody,
           nhdhr.id,
           #lagoslakeid,
           state_id) %>% 
  print(n = nrow(.))
```

#covreage
```{r}
library(tidyverse)
mn_lakes_ag <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Aggregated Data", "mn_lakes_hg_covariates.csv"))

mn_lakes_ag %>% 
  summarise(na_mglp = sum(is.na(nws_nlcd_wetemerg95_pct)),
            percent = (na_mglp/n()) * 100)

mn_lakes_ag %>% 
  distinct(state_id, nws_nlcd_wetemerg95_pct) %>% 
  summarise(na_mglp = sum(is.na(nws_nlcd_wetemerg95_pct)),
            percent = (na_mglp/n()) * 100)

mn_lakes_ag %>% 
  filter(is.na(nws_nlcd_wetemerg95_pct)) %>% 
  distinct(waterbody) %>% 
  arrange(waterbody) %>% 
  print(n = nrow(.))
```

#wisconsin data ag
```{r}
library(janitor)

#raw data
wi_1 <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "WI", "WI raw data from agency", "Statewide Fish Hg 1990-1999 for MN.csv"))

wi_2 <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "WI", "WI raw data from agency", "Statewide Fish Hg 2000-2009 for MN.csv"))

wi_3 <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "WI", "WI raw data from agency", "Statewide Fish Hg 2010-2017 for MN.csv"))

wi_4 <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "WI", "WI raw data from agency", "Statewide Fish Hg 2018-2023 for MN.csv"))

#aggregating
wi_hg <- bind_rows(wi_1, wi_2, wi_3, wi_4) %>% 
  clean_names() %>% 
  select(location_code:qualifier) %>% 
  mutate(sample_date = mdy(sample_date))
#write_csv(wi_hg, "wi_hg_26NOV2024.csv")
wi_hg <- read_csv(file.path("G:", 
                             "Shared drives", 
                             "Hansen Lab", 
                             "RESEARCH PROJECTS", 
                             "Statewide mercury modeling", 
                             "Data", 
                             "Mercury Data", 
                             "WI",
                             "wi_hg_26NOV2024.csv"))

glimpse(wi_hg)

#exploring the data
wi_hg %>% 
  group_by(location_type) %>% 
  count()
#almost 11K lake records

wi_hg %>% 
  filter(location_type == "LAKE") %>% 
  group_by(sample_type) %>% 
  count()

wi_hg %>% 
  filter(location_type == "LAKE") %>% 
  summarise(min_year = min(year(sample_date)),
            max_year = max(year(sample_date)))

wi_hg %>% 
  filter(location_type == "LAKE") %>% 
  summarise(n_distinct(wbic))

wi_hg %>% 
  filter(location_type == "LAKE") %>% 
  group_by(wbic, sample_type, year(sample_date)) %>% 
  count() %>% 
  group_by(sample_type) %>% 
  summarise(avg_n_spec_lake = mean(n),
            max = max(n),
            min = min(n))

wi_hg %>% 
  filter(location_type == "LAKE") %>% 
  group_by(wbic, sample_type) %>% 
  summarise(n_years = n_distinct(year(sample_date))) %>% 
  ggplot() +
  geom_histogram(aes(n_years)) +
  facet_wrap(~sample_type, scales = "free")

#wi zm lakes 
wi_zm <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Covariate Data", "WI_ZM_lakes.csv")) %>% 
  janitor::clean_names() %>% 
  mutate(wbic_clean = str_pad(waterbody_id_code_wbic, width = 7, side = "left", pad = "0")) %>% 
  select(species,
         status,
         wbic_clean,
         year_first_found)
glimpse(wi_zm)

wi_hg %>% 
  mutate(wbic_clean = str_replace(wbic, "_", "")) %>% 
  select(wbic, wbic_clean) %>% 
  distinct(wbic, wbic_clean) %>% 
  print(n = nrow(.))

wi_hg_zm <- wi_hg %>% 
  filter(location_type == "LAKE") %>% 
  mutate(wbic_clean = str_replace(wbic, "_", "")) %>%
  left_join(wi_zm, by = "wbic_clean", relationship = "many-to-one") %>% 
  mutate(zm_lake = case_when(!is.na(species) ~ "y",
                             TRUE ~ "n"),
         zm_sample = case_when(year_first_found >= year(sample_date) ~ "y",
                               TRUE ~ "n"),
         zm_sample_lag = case_when(year_first_found >= (year(sample_date) + 2) ~ "y",
                               TRUE ~ "n"))
wi_hg_zm %>% 
  distinct(wbic, zm_lake) %>% 
  summarise(zm_lakes = sum(zm_lake == "y"),
            uninvaded = sum(zm_lake == "n"))

wi_hg_zm %>% 
  group_by(zm_lake) %>% 
  count()

wi_hg_zm %>% 
  filter(zm_lake == "y") %>% 
  group_by(wbic_clean) %>% 
  count() %>% 
  ungroup() %>% 
  summarise(avg_invaded = mean(n))

wi_hg_zm %>% 
  filter(zm_lake == "y") %>% 
  distinct(wbic_clean, zm_sample) %>% 
  group_by(wbic_clean) %>% 
  count() %>% 
  filter(n >1)

wi_hg_zm %>% 
  filter(zm_lake == "y") %>% 
  group_by(wbic_clean, zm_sample) %>% 
  count() %>% 
  group_by(zm_sample) %>% 
  summarise(invaded_avg = mean(n),
            max_n = max(n),
            min_n = min(n))
```

# 2025 continuation data exploration
```{r}
zm_ssw <- read_csv("Data/zm_ssw_2025.csv") %>% 
  filter(!is.na(dow)) %>% 
  filter(!(dow %in% c("na", "none"))) %>% 
  distinct()
hg <- read_csv("Data/allfish_data_04042024_JLO.csv") 

#fix invasion data
clean_lake_ids <- function(id) {
  if (is.na(id) || id == "") return(NA_character_)  # Handle NA or empty values
  
  id_clean <- as.character(id)  # Ensure it's a character string
  id_clean <- str_replace_all(id_clean, "-", "")  # Remove dashes
  
  # Ensure it has a leading zero if it originally had one
  if (nchar(id_clean) < 6) {
    id_clean <- str_pad(id_clean, width = 6, pad = "0", side = "left")
  }
  
  # Ensure it's 8 digits long
  if (nchar(id_clean) == 6) {
    id_clean <- paste0(id_clean, "00")  # Append "00" if only 6 digits long
  }
  
  return(id_clean)
}


zm_ssw <- zm_ssw %>% 
  mutate(dow = sapply(dow, clean_lake_ids))
rm(clean_lake_ids)

zm_ssw <- zm_ssw %>% 
  select(dow, ais, year_infested) %>% 
  pivot_wider(names_from = ais,
              values_from = year_infested) %>% 
  rename(zm_year_infested = `zebra mussel`,
         ssw_year_infested = `starry stonewort`)

clean_dowid <- function(id) {
  if (is.na(id) || id == "") return(NA_character_)  # Handle NA values
  id_clean <- str_replace_all(id, "-", "")  # Remove dashes
  return(id_clean)
}

hg <- hg %>% 
  mutate(dow = map_chr(DOWID, clean_dowid)) %>% 
  mutate(dow = case_when(dow == "56074701" | dow == "56074702" ~ "56074700",
                         dow == "73020000" | dow == "73020002" ~ "73020000",
                         TRUE ~ dow)) 
rm(clean_dowid)

#zm and starry lake hg
zm_ssw_hg <- zm_ssw %>% 
  inner_join(hg)

ais_lakes <- zm_ssw_hg %>% 
  distinct(dow, YEARCOLL, SPEC, zm_year_infested, ssw_year_infested) %>% 
  mutate(pre_ssw = case_when(YEARCOLL < ssw_year_infested ~ "y",
                             TRUE ~ "n"),
         post_ssw = case_when(YEARCOLL > ssw_year_infested ~ "y",
                             TRUE ~ "n"),
         pre_zm = case_when(YEARCOLL < zm_year_infested ~ "y",
                             TRUE ~ "n"),
         post_zm = case_when(YEARCOLL > zm_year_infested ~ "y",
                             TRUE ~ "n"))
write_csv(ais_lakes, "dnr_hg_invasion.csv")

############################################################
#getting study lakes
study_dows <- c(49007900, 47004901, 3057600, 29018500, 1006200, 21014500, 86025200,
          34007900, 11030500, 11035800, 3015300, 73020000, 11016700, 3047500,
          56074700, 18030800, 49012700, 11050400, 56061300, 47004600, 11020300, 04003000, 11014700, 86009000)
study_dows <- str_pad(as.character(study_dows), width = 8, side = "left", pad = "0")
study_dows <- tibble(dow = study_dows)

study_comb <- study_dows %>% 
  left_join(zm_ssw, by = "dow") %>% 
  filter(!is.na(year_infested)) %>% 
  pivot_wider(id_cols = c(dow, lake_name),
              values_from = c(year_infested),
              names_from = ais) %>% 
  mutate(zm_invaded = case_when(!is.na(`zebra mussel`) ~ "y",
                                TRUE ~ "n"),
         ssw_invaded = case_when(!is.na(`starry stonewort`) ~ "y",
                                TRUE ~ "n"),
         both_invaded = case_when(!is.na(`zebra mussel`) & !is.na(`starry stonewort`) ~ "y",
                                TRUE ~ "n")) 

#nice, lets looks at some mercury values now
clean_dowid <- function(id) {
  if (is.na(id) || id == "") return(NA_character_)  # Handle NA values
  id_clean <- str_replace_all(id, "-", "")  # Remove dashes
  return(id_clean)
}

hg <- hg %>% 
  mutate(dow = map_chr(DOWID, clean_dowid)) %>% 
  mutate(dow = case_when(dow == "56074701" | dow == "56074702" ~ "56074700",
                         dow == "73020000" | dow == "73020002" ~ "73020000",
                         TRUE ~ dow)) 
rm(clean_dowid)

study_hg <- hg %>% 
  right_join(study_comb, "dow")

study_hg %>% 
  group_by(dow, lake_name) %>%
  summarise(n_hg = sum(!is.na(HGPPM)))

#how many samples pre invasion?
study_hg %>% 
  group_by(dow, lake_name, `zebra mussel`) %>%
  summarise(n_hg_pre = sum(YEARCOLL < `zebra mussel`),
            n_hg_post = sum(YEARCOLL >= `zebra mussel`))

study_hg %>% 
  group_by(dow, lake_name, `starry stonewort`) %>%
  summarise(n_hg_pre = sum(YEARCOLL < `starry stonewort`),
            n_hg_post = sum(YEARCOLL >= `starry stonewort`))

study_hg %>% 
  filter(dow == "03057600") %>% 
  mutate(zm = case_when(YEARCOLL < `zebra mussel` ~ "pre",
                        YEARCOLL >= `zebra mussel` ~ "post")) %>% 
  ggplot() +
  geom_point(aes(LGTHIN, HGPPM, color = zm)) +
  geom_smooth(aes(LGTHIN, HGPPM, color = zm)) +
  facet_wrap(~SPEC)

#any non-zm but ssw lakes?
zm_ssw_lakes <- zm_ssw %>% 
  filter(!is.na(dow)) %>%
  filter(dow != "21021200") %>% 
  pivot_wider(id_cols = c(dow, lake_name),
              values_from = c(year_infested),
              names_from = ais)
zm_ssw_lakes %>% 
  filter(is.na(`zebra mussel` & !is.na(`starry stonewort`)))
#upper red, wolf, and rush actually have zm but they list zm with a different dow in list

ssw_lakes <- zm_ssw_lakes %>% 
  filter(is.na(`zebra mussel` & !is.na(`starry stonewort`))) %>% 
  filter(!(dow %in% c("04007900", "04003501", "18031100"))) %>% 
  mutate(dow)

ssw_hg <- hg %>% 
  right_join(ssw_lakes)
```

#can we assume overlap of MN and WI state data?
```{r}
wi_hg <- read_csv(file.path("G:", 
                             "Shared drives", 
                             "Hansen Lab", 
                             "RESEARCH PROJECTS", 
                             "Statewide mercury modeling", 
                             "Data", 
                             "Mercury Data", 
                             "WI",
                             "wi_hg_26NOV2024.csv"))

mn_dnr <- read_csv(file.path("G:",
                             "Shared drives",
                             "Hansen Lab",
                             "RESEARCH PROJECTS",
                             "Statewide mercury modeling",
                             "Data",
                             "Mercury Data",
                             "allfish_data_04042024_JLO.csv"))

glifwc <- read_csv(file.path("G:",
                             "Shared drives",
                             "Hansen Lab",
                             "RESEARCH PROJECTS",
                             "Statewide mercury modeling",
                             "Data",
                             "Mercury Data",
                             "glifwc_hg.csv"))

###############################################################################
glifwc %>% 
  group_by(state) %>% 
  count()

glifwc %>% 
  filter(state == "MN") %>% 
  group_by(waterbody, year(sample_date)) %>% 
  count() %>% 
  print(n = nrow(.))

mn_dnr %>% 
  filter(OWNER == "GLIFWC") %>% 
  filter(WATERWAY == "MILLE LACS" & SPEC == "WAE") %>% 
  group_by(YEARCOLL) %>% 
  count() %>% 
  print(n = nrow(.))

glifwc %>% 
  filter(str_detect("MILLE LACS L", waterbody)) %>% 
  group_by(year(sample_date)) %>% 
  count() %>% 
  print(n = nrow(.))
#looks like the glifwc data is more complete than what the state has 

mn_dnr %>% 
  filter(OWNER == "GLIFWC") %>% 
  group_by(WATERWAY) %>% 
  count()

glifwc %>% 
  filter(state == "MN") %>% 
  group_by(waterbody) %>% 
  count() %>% 
  print(n = nrow(.))

wi_hg %>% 
  filter(collector == "GLIFWC") %>% 
  group_by(wmu_name) %>% 
  count()

glifwc %>% 
  filter(state == "WI") %>% 
  group_by(waterbody) %>% 
  count() %>% 
  print(n = nrow(.))
```


# exploring expanded data
```{r}
glifwc <- read_csv(file.path("G:", 
                             "Shared drives", 
                             "Hansen Lab", 
                             "RESEARCH PROJECTS", 
                             "Statewide mercury modeling", 
                             "Data", "Mercury Data", 
                             "glifwc_hg.csv"))

glifwc %>% 
  group_by(spp_code) %>% 
  summarise(min_year = min(year),
            max_year = max(year))

glifwc %>% 
  summarise(n_lakes = n_distinct(waterbody))
rm(glifwc)

wi <- read_csv(file.path("G:", 
                             "Shared drives", 
                             "Hansen Lab", 
                             "RESEARCH PROJECTS", 
                             "Statewide mercury modeling", 
                             "Data", 
                             "Mercury Data", 
                             "WI",
                             "wi_hg_26NOV2024.csv"))

wi %>% 
  mutate(year = year(sample_date)) %>% 
  group_by(sample_type) %>% 
  summarise(min = min(year),
            max = max(year),
            n = n())

wi %>% 
  summarise(n_lakes = n_distinct(wbic))

wi %>% 
  group_by(location_type) %>% 
  count()


```

# pulling lake info for MASIC for BACI analysis - data came the clean BACI script
```{r}
library(arrow)
lakes <- combined_data %>% 
  distinct(lake_name, lake_id, centroid_lat, centroid_lon, year_infested, zm_lake)

mw_county <- open_dataset(file.path("G:",
                                    "Shared drives",
                                    "Hansen Lab",
                                    "RESEARCH PROJECTS",
                                    "Fish Survey Data",
                                    "Parquet Files",
                                    "hive_update"),
                          partitioning = "state") %>% 
  filter(state == "Minnesota")

mw_county <- mw_county %>% 
  filter(!is.na(lake_id)) %>% 
  filter(!is.na(county)) %>% 
  distinct(lake_id, county) %>% 
  collect()

mw_county <- mw_county %>% 
  mutate(lake_id = as.character(lake_id),
         lake_id = case_when(nchar(lake_id) == 7 ~ paste0("0", lake_id),
                             TRUE ~ lake_id))

lakes <- lakes %>% 
  left_join(mw_county)

lakes <- lakes %>% 
  mutate(county = case_when(lake_id == "70012000" ~ "Scott",
                            lake_id == "47004900" ~ "Meeker",
                            lake_id == "56076000" ~ "Otter Tail",
                            TRUE ~ county))
write_csv(lakes,"zm_hg_lakes.csv")
```