---
title: "zm_baci_update"
author: "Denver Link"
date: "2024-12-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Library
```{r}
library(tidyverse)
library(mwlaxeref)
library(janitor)
library(mnsentinellakes)
library(sf)
library(brms)
library(emmeans)
library(tidybayes)
```

# Data
```{r}
#mercury data
hg <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Mercury Data", "allfish_data_04042024_JLO.csv")) %>% 
  filter(!(is.na(LGTHIN) | is.na(HGPPM) | is.na(DOWID))) %>%   #removes samples with no hg, length, or dow data
  filter(TYPE %in% c("Lake", "LAKE")) %>% 
  mutate(DOWID = fixlakeid(DOWID)) %>% 
  filter(DOWID != "16000100") %>% 
  local_to_nhdhr(from_colname = "DOWID", states = "mn") %>% 
  mutate(nhdhr.id = str_remove(nhdhr.id, "nhdhr_")) %>% 
  select(SAMPLENO,
         WATERWAY,
         TYPE,
         LOCATION,
         DOWID,
         nhdhr.id,
         DATECOL2,
         YEARCOLL,
         SPEC,
         ANATHG,
         NOFISH,
         LGTHIN,
         WTLB,
         AGE,
         SEX,
         HGPPM,
         HGCODE,
         HGLAB,
         HGCMMT) %>% 
  rename(DOW = DOWID,
         nhdid = nhdhr.id) %>% 
  ungroup()

#connection to covariates
dnr_covary <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Covariate Data", "MN_lakes_watershed_lulc_fromDNR.csv")) %>% 
  mutate(logit_wetlands = car::logit(Percent_Wetland),
         logit_urban = car::logit(Percent_Urban),
         logit_ag = car::logit(Percent_Cultivated_Crops),
         log_WS_Lake_Ratio = log(WS_Lake_Ratio))

#scaled_dnr <- dnr_covary %>%
  #mutate_at(vars(Total_Pasture_Hay_Hectares:log_WS_Lake_Ratio), scale) %>% 
  #rename_with(~ paste0("scaled_", .), Total_Pasture_Hay_Hectares:log_WS_Lake_Ratio) 

#dnr_covary <- dnr_covary %>% 
  #left_join(scaled_dnr)

#adding dnr covariates into the data
hg <- hg %>% 
  left_join(dnr_covary)
rm(dnr_covary, scaled_dnr)

#glm - for clarity and area values
glm <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Data", "Covariate Data", "glm_lake_metadata.csv")) %>%
  filter(state == "MN") %>% 
  rename(nhdid = site_id) %>% 
  mutate(nhdid = gsub("^nhdhr_", "", nhdid)) %>% 
  mutate(clarity = case_when(is.infinite(clarity) ~ NA,
                             TRUE ~ clarity)) %>% 
  mutate(log_area = log(area))


hg <- hg %>% 
  left_join(glm) %>% 
  mutate(across(starts_with("scaled_"), as.vector))
rm(glm)
glimpse(hg)

#infested waters list
zm <- read_csv("Data/ZM_MN_NHD.csv") %>% 
  filter(!is.na(DOW)) %>% 
  mutate(DOW = as.character(gsub("mndow_", "", MNDOW_ID)),
         nhdid = NHD_ID) %>% 
  select(DOW,
         Year_Infested) %>% 
  mutate(zm_lag_year_infested = Year_Infested +2)

hg <- hg %>% 
  left_join(zm, by = c("DOW"))
rm(zm)

#lake link
lake_link <- read_csv("Data/lake_link.csv") %>% 
  distinct(lake_nhdid, .keep_all = T) %>% 
  rename(nhdid = lake_nhdid)

lake_link_sf <- lake_link %>% 
  st_as_sf(coords = c("lake_lon_decdeg", "lake_lat_decdeg"), crs = 4326) %>% 
  st_transform(crs = 32615)
utm_coords <- st_coordinates(lake_link_sf)
lake_link <- lake_link %>%
  mutate(easting = utm_coords[, 1]/1000,
         northing = utm_coords[, 2]/1000)
#utm values in km instead of m
rm(lake_link_sf, utm_coords)

hg <- hg %>% 
  left_join(lake_link)
rm(lake_link)


#data for modeling - select for columns of use 
hg.model <- hg %>%
  select(WATERWAY,
         TYPE,
         DOW,
         nhdid,
         DATECOL2,
         YEARCOLL,
         SPEC,
         ANATHG,
         NOFISH,
         LGTHIN,
         HGPPM,
         Total_Wetland_Hectares,
         Percent_Wetland,
         Percent_Urban,
         Percent_Cultivated_Crops,
         logit_wetlands,
         logit_urban,
         logit_ag,
         WS_Lake_Ratio,
         log_WS_Lake_Ratio,
         centroid_lat,
         centroid_lon,
         max_depth,
         area,
         log_area,
         elevation,
         clarity,
         Year_Infested,
         zm_lag_year_infested,
         lake_lat_decdeg,
         lake_lon_decdeg,
         easting,
         northing) %>% 
  mutate(zm_lake = case_when(is.na(Year_Infested) ~ "n",
                             TRUE ~ "y"),
         zm_sample = case_when(zm_lag_year_infested <= YEARCOLL ~ "y",
                               TRUE ~ "n"),
         time_since_invasion = YEARCOLL - zm_lag_year_infested) %>% 
  clean_names()

rm(hg)
```

# Data Coverage
```{r}
# zebra mussel coverage
hg.model %>% 
  group_by(zm_lake) %>% 
  summarise(total_samples = n(),
            pre_zm = sum(zm_sample == "n"),
            post_zm = sum(zm_sample == "y"),
            distinct_lakes = n_distinct(dow),
            distinct_years = n_distinct(yearcoll))

#how many zm samples and lakes in 2023?
hg.model %>% 
  filter(zm_lake == "y" & yearcoll == "2023") %>% 
  summarise(zm_sampling_2023 = n(),
            zm_sampling_lakes_2023 = n_distinct(dow)
  )

#how many zebra mussel lakes have post data - how many years of post?
hg.model %>% 
  filter(zm_lake == "y") %>% 
  distinct(dow, zm_sample) %>% 
  group_by(dow) %>% 
  count() %>% 
  filter(n > 1)

hg.model %>% 
  filter(zm_lake == "y" & zm_sample == "y") %>% 
  group_by(dow) %>% 
  summarise(n_years = n_distinct(yearcoll)) %>% 
  ggplot() +
  geom_histogram(aes(x = n_years)) +
  geom_histogram(data = . %>% filter(dow == "48000200"), aes(n_years, fill = "red"))+
  theme(legend.position = "none")

#for lakes sampled in 2023 - how many years of post data do they have?
lakes_2023 <- hg.model %>% 
  filter(yearcoll == 2023 & zm_lake == "y") %>% 
  distinct(dow, waterway)

hg.model %>% 
  filter(dow %in% c(lakes_2023$dow)) %>% 
  filter(zm_sample == "y") %>% 
  group_by(dow) %>% 
  summarise(n_years = n_distinct(yearcoll)) %>% 
  ggplot() +
  geom_histogram(aes(n_years))
# 10/18 zebra mussel lake sampled in 2023 were sampled for the first time
rm(lakes_2023)
```

# Data for Analysis 
```{r}
#walleye zebra mussel data 
wae_zm_lakes <- hg.model %>% 
  filter(spec == "WAE") %>% 
  filter(yearcoll >= 2000) %>% 
  filter(zm_lake == "y") %>% 
  distinct(dow, zm_sample) %>% 
  group_by(dow) %>% 
  count() %>% 
  filter(n >1)
#this code chunk 1.) filters for walleye 2.) filters for years 2000-most recent 3.) filters for lakes invaded by zebra mussels 4.) finds the unique combinations for lake id and hg samples collected during zm invasion 5.) counts how many times a lake shows up - if it shows up 2x then it has both y and n for zm invasion samples 6.) filters for lakes that only have zm invasion 

wae_zm_fish <- hg.model %>% 
  filter(yearcoll >= 2000) %>%
  filter(spec == "WAE") %>% 
  filter(dow %in% c(wae_zm_lakes$dow))

#walleye reference lake 
wae_ref_lakes <- hg.model %>% 
  filter(spec == "WAE") %>% 
  filter(yearcoll >= 2000) %>% 
  filter(zm_lake == "n") %>% 
  #creating pseudo invasion based on middle year of the study
  mutate(zm_sample = case_when(yearcoll >= 2011 ~ "y",
                               yearcoll <= 2010 ~ "n")) %>% 
  distinct(dow, zm_sample) %>% 
  group_by(dow) %>% 
  count() %>% 
  filter(n >1)

wae_ref_fish <- hg.model %>% 
  filter(yearcoll >= 2000) %>%
  filter(spec == "WAE") %>% 
  filter(dow %in% c(wae_ref_lakes$dow)) %>% 
  mutate(zm_sample = case_when(yearcoll >= 2011 ~ "y",
                               yearcoll <= 2010 ~ "n"))

#######################northern pike#######################################
#pike zebra mussel data 
nop_zm_lakes <- hg.model %>% 
  filter(spec == "NOP") %>% 
  filter(yearcoll >= 2000) %>% 
  filter(zm_lake == "y") %>% 
  distinct(dow, zm_sample) %>% 
  group_by(dow) %>% 
  count() %>% 
  filter(n >1)
#this code chunk 1.) filters for pike 2.) filters for years 2000-most recent 3.) filters for lakes invaded by zebra mussels 4.) finds the unique combinations for lake id and hg samples collected during zm invasion 5.) counts how many times a lake shows up - if it shows up 2x then it has both y and n for zm invasion samples 6.) filters for lakes that only have zm invasion 

nop_zm_fish <- hg.model %>% 
  filter(yearcoll >= 2000) %>%
  filter(spec == "NOP") %>% 
  filter(dow %in% c(nop_zm_lakes$dow))

#walleye reference lake 
nop_ref_lakes <- hg.model %>% 
  filter(spec == "NOP") %>% 
  filter(yearcoll >= 2000) %>% 
  filter(zm_lake == "n") %>% 
  #creating pseudo invasion based on middle year of the study
  mutate(zm_sample = case_when(yearcoll >= 2011 ~ "y",
                               yearcoll <= 2010 ~ "n")) %>% 
  distinct(dow, zm_sample) %>% 
  group_by(dow) %>% 
  count() %>% 
  filter(n >1)

nop_ref_fish <- hg.model %>% 
  filter(yearcoll >= 2000) %>% 
  filter(spec == "NOP") %>% 
  filter(dow %in% c(nop_ref_lakes$dow))  %>% 
  mutate(zm_sample = case_when(yearcoll >= 2011 ~ "y",
                               yearcoll <= 2010 ~ "n"))

##############combined####################
baci_data <- bind_rows(nop_ref_fish, nop_zm_fish, wae_ref_fish, wae_zm_fish)
rm(nop_ref_fish, nop_zm_fish, wae_ref_fish, wae_zm_fish,
   nop_ref_lakes, nop_zm_lakes, wae_ref_lakes, wae_zm_lakes)

baci_data <- baci_data %>% 
  group_by(spec) %>% 
  mutate(scaled.length = c(scale(lgthin)))

###########quick QC##################################
baci_data %>% 
  group_by(yearcoll) %>% 
  count() %>% 
  print(n = nrow(.))

baci_data %>% 
  group_by(zm_lake, zm_sample) %>% 
  count()
```
Data included:
1. species = walleye and northern pike
2. years = 2000 - 2023
3. zm lakes with at least one year pre and post invasion (invasion considered without year lag) *this is most likley the difference between denver's master's analysis in sample sizes*
4. reference lakes included with at least one year before and after 2011 (pseudo invasion 2011 - mid point of years in the analysis )

# Analysis 
```{r}
fit_1 <- brm(log(hgppm) ~ scaled.length*spec + zm_lake*zm_sample*spec + (1|dow) + (1|yearcoll), 
             data = baci_data, 
             family = gaussian(), 
             iter = 10000)
#saveRDS(fit_1, "baci_update.rds")
fit_1 <- readRDS("baci_update.rds")
summary(fit_1)
conditional_effects(fit_1)

#look at the did package to fit a model that includes time since invasion vs just a pre-post
```

# Post Analysis - fit 1
```{r}
marg_effect <- emmeans(fit_1, ~ zm_lake * zm_sample | spec, re_fromula = NA, epred = T)
contrast(marge_effect)

#marginal plot - concider adding epred = T and no random effects (not sure how the calls work with this function)
marg_plot <- emmip(fit_1, zm_lake ~ zm_sample | spec, CIs = T)
marg_plot +
  scale_color_manual("Zebra Mussel Lake", labels = c("No", "Yes"), values = c("lightblue", "salmon"))

#pulling out the difference between groups 
marg_draws <- emmeans(fit_1, ~ zm_lake * zm_sample | spec, re_fromula = NA, epred = T) %>% 
  contrast() %>% 
  gather_emmeans_draws() %>% 
  unite("contrast_spec", contrast, spec, sep = "_") %>% # Combine contrast and spec into one column
  pivot_wider(
    names_from = contrast_spec, # Use the combined column for new column names
    values_from = .value        # Populate new columns with .value
  ) %>% 
  clean_names() %>% 
  mutate(ref_diff_wae = n_y_effect_wae - n_n_effect_wae,
         zm_diff_wae = y_y_effect_wae - y_n_effect_wae,
         ref_diff_nop = n_y_effect_nop - n_n_effect_nop,
         zm_diff_nop = y_y_effect_nop - y_n_effect_nop) %>% 
  select(ref_diff_wae,
         zm_diff_wae,
         ref_diff_nop,
         zm_diff_nop) %>% 
  pivot_longer(
    cols = c(ref_diff_wae, zm_diff_wae, ref_diff_nop, zm_diff_nop),
    names_to = c("zm_lake", "spec"),
    names_pattern = "(ref|zm)_diff_(.*)",
    values_to = "diff"
  )

marg_draws %>% 
  ggplot() +
  geom_density(aes(x = diff, fill = zm_lake), alpha = .5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~spec) +
  scale_fill_manual("Zebra Mussle Invasion", labels = c("No", "Yes"), values = c("lightblue", "salmon")) +
  theme_minimal() +
  theme(legend.position = "bottom")

#diff of differences 
diff_of_diff <- emmeans(fit_1, ~ zm_lake * zm_sample | spec, re_fromula = NA, epred = T) %>% 
  contrast() %>% 
  gather_emmeans_draws() %>% 
  unite("contrast_spec", contrast, spec, sep = "_") %>% # Combine contrast and spec into one column
  pivot_wider(
    names_from = contrast_spec, # Use the combined column for new column names
    values_from = .value        # Populate new columns with .value
  ) %>% 
  clean_names() %>% 
  mutate(ref_diff_wae = n_y_effect_wae - n_n_effect_wae,
         zm_diff_wae = y_y_effect_wae - y_n_effect_wae,
         ref_diff_nop = n_y_effect_nop - n_n_effect_nop,
         zm_diff_nop = y_y_effect_nop - y_n_effect_nop) %>% 
  select(ref_diff_wae,
         zm_diff_wae,
         ref_diff_nop,
         zm_diff_nop) %>%  
  mutate(diff_of_diff_wae = zm_diff_wae - ref_diff_wae,
         diff_of_diff_nop = zm_diff_nop - ref_diff_nop) %>% 
   pivot_longer(
    cols = starts_with("diff_of_diff"), # Select columns with 'diff_of_diff'
    names_to = "species",               # New column for the species names
    names_prefix = "diff_of_diff_",     # Remove this prefix from the names
    values_to = "value"                 # New column for the values
  ) %>% 
  select(species,
         value)

diff_of_diff_sum <- diff_of_diff %>% 
  group_by(species) %>% 
  summarise(median_diff = median(value),
             lower_66 = quantile(value, probs = 0.17), # Lower bound of 66% credible interval
    upper_66 = quantile(value, probs = 0.83), # Upper bound of 66% credible interval
    lower_95 = quantile(value, probs = 0.025), # Lower bound of 95% credible interval
    upper_95 = quantile(value, probs = 0.975)  # Upper bound of 95% credible interval
            )

diff_of_diff %>% 
  ggplot() +
  geom_density(aes(x = value)) + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~species)

#linewidth does not work in making the 66th visable - work on that 
diff_of_diff_sum %>% 
  ggplot() +
  geom_pointinterval(aes(x = species, y = median_diff, ymin = lower_95, ymax = upper_95), linewidth = .5) +
  geom_pointinterval(aes(x = species, y = median_diff, ymin = lower_66, ymax = upper_66), linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed")
```

