---
title: "zm_baci_update_clean"
author: "Denver Link"
date: "2025-03-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Library
```{r}
library(tidyverse)
library(mwlaxeref)
library(janitor)
library(mnsentinellakes)
library(sf)
library(brms)
library(emmeans)
library(tidybayes)
library(patchwork)
library(cowplot)
library(ggpubr)
library(sf)
library(tigris)
library(ggspatial)
library(gridExtra)
library(ggimage)


#a function from mnsentinel lakes
fixlakeid <- function(lakeid){
  lakeid = gsub("-", "", as.character(lakeid))
  lakeid[nchar(lakeid) == 7] = paste0("0", lakeid[nchar(lakeid) == 7]) # this line specifies to paste a leading zero only onto those values with 7 characters
  lakeid # r will, by default, return the last variable in the body
}

```

# Data - already filtered 
*data below the LOD was already cut in half*
```{r}
combined_data <- read_csv(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "combined_hg_data.csv"))

#break out for each species
wae_data <- combined_data %>% 
  filter(spec == "WAE")

nop_data <- combined_data %>% 
  filter(spec == "NOP")

blc_data <- combined_data %>% 
  filter(spec == "BLC")

blg_data <- combined_data %>% 
  filter(spec == "BLC")

lmb_data <- combined_data %>% 
  filter(spec == "LMB")

yep_data <- combined_data %>% 
  filter(spec == "YEP")
```

###########Model fitting##################

#model runs - each individual species run seperatly
```{r}
#set priors for all models so they are not informed by they model 
priors <- c(prior(normal(1,100), class = Intercept),
            prior(student_t(3, 0, 2.5), class = sd),
            prior(student_t(3,0,2.5), class = sigma))

#walleye model fit
fit_wae <- brm(log(hgppm) ~ scaled_length + zm_lake*zm_sample_lag_2 + (1|dow) + (1|yearcoll), 
             data = wae_data, 
             family = gaussian(), 
             iter = 20000,
             warmup = 5000,
             prior = priors)
#saveRDS(fit_wae, "baci_update_wae.rds")

#northern pike model fit
fit_nop <- brm(log(hgppm) ~ scaled_length + zm_lake*zm_sample_lag_2 + (1|dow) + (1|yearcoll), 
             data = nop_data, 
             family = gaussian(), 
             iter = 20000,
             warmup = 5000,
             prior = priors)
#saveRDS(fit_nop, "baci_update_nop.rds")

fit_blg <- brm(log(hgppm) ~ scaled_length + zm_lake*zm_sample_lag_2 + (1|dow) + (1|yearcoll), 
             data = blg_data, 
             family = gaussian(), 
             iter = 20000,
             warmup = 5000,
             prior = priors)
#saveRDS(fit_blg, "baci_update_blg.rds")

fit_blc <- brm(log(hgppm) ~ scaled_length + zm_lake*zm_sample_lag_2 + (1|dow) + (1|yearcoll), 
             data = blc_data, 
             family = gaussian(), 
             iter = 20000,
             warmup = 5000,
             prior = priors)
#saveRDS(fit_blc, "baci_update_blc.rds")

fit_lmb <- brm(log(hgppm) ~ scaled_length + zm_lake*zm_sample_lag_2 + (1|dow) + (1|yearcoll), 
             data = lmb_data, 
             family = gaussian(), 
             iter = 20000,
             warmup = 5000,
             prior = priors)
#saveRDS(fit_lmb, "baci_update_lmb.rds")

fit_yep <- brm(log(hgppm) ~ scaled_length + zm_lake*zm_sample_lag_2 + (1|dow) + (1|yearcoll), 
             data = yep_data, 
             family = gaussian(), 
             iter = 20000,
             warmup = 5000,
             prior = priors)
#saveRDS(fit_yep, "baci_update_yep.rds")
```

#saved model runs
```{r}
fit_wae <- readRDS(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Output", "baci_update_wae.rds"))
fit_nop <- readRDS(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Output", "baci_update_nop.rds"))
fit_blg <- readRDS(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Output", "baci_update_blg.rds"))
fit_lmb <- readRDS(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Output", "baci_update_lmb.rds"))
fit_yep <- readRDS(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Output", "baci_update_yep.rds"))
fit_blc <- readRDS(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Output", "baci_update_blc.rds"))
```

######### Analysis #####################

#basic model fitting summaries 
```{r}
#priors
prior_summary(fit_wae)
prior_summary(fit_nop)
prior_summary(fit_blg)
prior_summary(fit_blc)
prior_summary(fit_lmb)
prior_summary(fit_yep)

#model output
summary(fit_wae)
summary(fit_nop)
summary(fit_blg)
summary(fit_blc)
summary(fit_lmb)
summary(fit_yep)

#plots
plot(fit_wae)
plot(fit_nop)
plot(fit_blg)
plot(fit_blc)
plot(fit_lmb)
plot(fit_yep)

#posterior checks
pp_check(fit_wae)
pp_check(fit_nop)
pp_check(fit_blg)
pp_check(fit_blc)
pp_check(fit_lmb)
pp_check(fit_yep)
```

# Pulling marginal effects for all species
```{r}
#create a species specific lengths for which the marginal effect will be calculated
species_lengths <- tibble(
  species = c("wae", "nop", "blc", "blg", "lmb", "yep"),
  standard_length = c(381, 610, 205, 155, 305, 205) # Adjust as needed
)

# Create a named list of datasets
species_datasets <- list(
  wae = wae_data,
  nop = nop_data,
  blc = blc_data,
  blg = blg_data,
  lmb = lmb_data,
  yep = yep_data
)

# Get summary stats for each species length to convert lengths to scaled lengths
species_summary_stats <- map(species_datasets, ~ summarise(.x, mean_length = mean(lgthmm), sd_length = sd(lgthmm)))

# Extract marginal effects and combine into one data frame
# specific for each species, give emmeans the model fit and length
# calculate expected marginal effect ignoring specific random effects 
# convert the log scale back value back to hg ppm
marginal_effects_all <- species_lengths %>%
  mutate(marginal_draws = pmap(list(species, standard_length), function(sp, len) {
    length_stats <- species_summary_stats[[sp]]
    specific_length_scaled <- (len - length_stats$mean_length) / length_stats$sd_length
    
    emmeans(get(paste0("fit_", sp)), ~ zm_lake * zm_sample_lag_2,
            at = list(scaled_length = specific_length_scaled),
            re_formula = NA, epred = TRUE) %>%
      gather_emmeans_draws() %>%
      mutate(species_name = sp, hg_ppm = exp(.value))
  })) %>%
  unnest(marginal_draws)

#we should see a marginal effect estimate for each draw from the postiror distribution for each species
# 15,000 draws X 4 chains X 6 species X 4 effects per draws = 1,440,000 obs
glimpse(marginal_effects_all)

# generate summaries for plotting

#first clean up the invasion naming scheme
marg_draw_plot <- marginal_effects_all %>%
      mutate(interaction = paste0(zm_lake, zm_sample_lag_2)) %>%
      mutate(interaction = case_when(
        interaction == "nn" ~ "Uninvaded Before",
        interaction == "ny" ~ "Uninvaded After",
        interaction == "yn" ~ "Zebra Mussel Before",
        interaction == "yy" ~ "Zebra Mussel After",
        TRUE ~ NA
      ),
      interaction = factor(interaction, levels = c("Uninvaded Before", "Uninvaded After", "Zebra Mussel Before", "Zebra Mussel After")))

# create lines connecting pre and post medians for each speciesXinvasion
median_lines_plot <- marg_draw_plot %>%
      group_by(interaction, zm_lake, species) %>%
      summarize(median_hg_ppm = median(hg_ppm), .groups = "drop") %>%
      mutate(zm_lake = case_when(zm_lake == "n" ~ "Uninvaded",
                                 zm_lake == "y" ~ "Zebra Mussel"),
             zm_sample_lag_2 = case_when(str_detect(interaction, "efore") ~ "Before",
                                         TRUE ~ "After")) %>% 
  mutate(species = as.factor(case_when(species == "wae" ~ "Walleye",
                             species == "nop" ~ "Northern Pike",
                             species == "lmb" ~ "Largemouth Bass",
                             species == "blc" ~ "Black Crappie",
                             species == "yep" ~ "Yellow Perch",
                             species == "blg" ~ "Bluegill",
                             TRUE ~ species)),
         species = fct_rev(factor(species, levels = c("Northern Pike",
                                                      "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")))) 

#compute differences pre and post for each speciesXinvasion
contrasts_draws <- marginal_effects_all %>% 
      select(-.chain, -.iteration, -standard_length, -species_name) %>%
  group_by(species, .draw) %>% 
      pivot_wider(names_from = c(zm_lake, zm_sample_lag_2),
                  values_from = c(.value, hg_ppm))
#compute the difference between zm lake and uninvaded lakes for each species
diff_of_diff <- contrasts_draws %>% 
    group_by(species) %>% 
      mutate(diff_reference = hg_ppm_n_y - hg_ppm_n_n,
             diff_zm = hg_ppm_y_y - hg_ppm_y_n)  %>% 
      select(species, .draw, diff_reference, diff_zm) %>% 
      mutate(value = diff_zm - diff_reference) %>% 
      select(species, .draw, value) %>% 
      mutate(zm_lake = "Difference in Change")

diff_lakes <- contrasts_draws %>% 
  group_by(species) %>% 
      mutate(diff_reference = hg_ppm_n_y - hg_ppm_n_n,
             diff_zm = hg_ppm_y_y - hg_ppm_y_n) %>% 
      select(species, .draw, diff_reference, diff_zm) %>% 
      pivot_longer(cols = c(diff_reference, diff_zm),
                   names_to = "zm_lake",
                   values_to = "value") %>% 
      mutate(zm_lake = case_when(zm_lake == "diff_zm" ~ "Zebra Mussel",
                                 TRUE ~ "Uninvaded"))
```

# marginal effects plots
```{r}
#fish icons
species_images <- data.frame(
  species = c("Walleye", 
              "Northern Pike", 
              "Largemouth Bass", 
              "Black Crappie",
              "Yellow Perch", 
              "Bluegill"),
  img = c(file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "walleye_outline.jpg"), 
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "pike_outline.jpg"), 
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "largemouth_bass_outline.jpg"), 
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "crappie_outline.jpg"),
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "perch_outline.jpg"),
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "bluegill_outline.jpg"))
)

diff_of_diff %>% 
  ggplot() +
  stat_pointinterval(aes(x = species, y = value), .width = c(.66, .95)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Difference of Before and After Impact \nBetween Zebra Mussel and Reference Lakes") +
  xlab("") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.title.y = element_text(size = 12)) 

plot1 <- diff_of_diff %>% 
  mutate(species = as.factor(case_when(species == "wae" ~ "Walleye",
                             species == "nop" ~ "Northern Pike",
                             species == "lmb" ~ "Largemouth Bass",
                             species == "blc" ~ "Black Crappie",
                             species == "yep" ~ "Yellow Perch",
                             species == "blg" ~ "Bluegill",
                             TRUE ~ species)),
         species = fct_rev(factor(species, levels = c("Northern Pike",
                                                      "Walleye", 
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")))) %>% 
  ggplot() +
  stat_pointinterval(aes(x = value, y = species), .width = c(.66, .95)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  xlab("Difference Before and After Impact \nBetween Zebra Mussel and Reference Lakes") +
  ylab("") +
  scale_x_continuous(limits = c(-0.04, 0.04))+
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.title.y = element_text(size = 12))
plot1

plot1 <- plot1 +
  geom_image(data = species_images, aes(y = species, x = -.035, image = img), size = .15)
plot1
#ggsave("diff_of_diff.jpg", width = 7, height = 5, dpi = 600)


plot2 <- marg_draw_plot %>%
  mutate(species = as.factor(case_when(species == "wae" ~ "Walleye",
                             species == "nop" ~ "Northern Pike",
                             species == "lmb" ~ "Largemouth Bass",
                             species == "blc" ~ "Black Crappie",
                             species == "yep" ~ "Yellow Perch",
                             species == "blg" ~ "Bluegill",
                             TRUE ~ species)),
         species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill"))) %>%
  mutate(zm_sample_lag_2 = as.factor(case_when(zm_sample_lag_2 == "n" ~ "Before",
                                     TRUE ~ "After")),
         zm_sample_lag_2 = factor(zm_sample_lag_2, levels = c("Before", "After"))) %>% 
  ggplot(aes(x = zm_sample_lag_2, y = hg_ppm, color = zm_lake, shape = zm_lake)) +
  facet_wrap(~species, scale = "free") +
  stat_pointinterval(
    .width = c(.66, .95),
    show.legend = F,
    position = position_dodge(width = 0.2)
  )  +
  scale_shape_manual("Sample Invasion",
                     values = c(17, 16)) +
  geom_line(
    data = median_lines_plot,
    aes(x = zm_sample_lag_2, y = median_hg_ppm, group = zm_lake),
    position = position_dodge(width = 0.1), # Align lines with intervals
    inherit.aes = FALSE,
    key_glyph = "rect",
    #color = c("lightblue", "lightblue", "salmon", "salmon")
  ) +
  geom_hline(yintercept = 0.22, linetype = "dashed") +
  ylab("Mercury Concentration (ppm)") +
  xlab("") +
   scale_color_manual("Lake Invasion", 
                      values = c("lightblue", "salmon")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.title.y = element_text(size = 10),
    strip.background = element_rect(color = "black", fill = NA, linewidth  = 0.5),
    strip.text = element_text(face = "bold", size = 6),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5) 
  ) 
plot2

#get species images and positions for the second plot
species_images <- marg_draw_plot %>% 
  group_by(species_name, zm_lake) %>%
  median_qi(hg_ppm, width = c(.95)) %>%
  filter(zm_lake == "n") %>% 
  select(species_name, hg_ppm.upper) %>% 
  group_by(species_name) %>% 
  summarise(y = case_when(species_name == "nop" ~ .34,
                          species_name == "wae" ~ .215,
                          species_name == "lmb" ~ .205,
                          species_name == "blc" ~ .2,
                          species_name == "yep" ~ .2,
                          species_name == "blg" ~ .2)) %>% 
  mutate(species = as.factor(case_when(species_name == "wae" ~ "Walleye",
                             species_name == "nop" ~ "Northern Pike",
                             species_name == "lmb" ~ "Largemouth Bass",
                             species_name == "blc" ~ "Black Crappie",
                             species_name == "yep" ~ "Yellow Perch",
                             species_name == "blg" ~ "Bluegill",
                             TRUE ~ species_name)),
         species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")),
         img = case_when(species == "Walleye" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "walleye_outline.jpg"),
                         species == "Northern Pike" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "pike_outline.jpg"),
                         species == "Largemouth Bass" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "largemouth_bass_outline.jpg"),
                         species == "Black Crappie" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "crappie_outline.jpg"),
                         species == "Yellow Perch" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "perch_outline.jpg"),
                         species == "Bluegill" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "bluegill_outline.jpg")),
         x = 1.5)

plot2_icons <- plot2 +
  geom_image(data = species_images, aes(y = y, x = x, image = img), size = .3, inherit.aes = F)
plot2_icons
#ggsave("baci_spec.jpg", width = 7, height = 5, dpi = 600)



#combined
combined_plot <-  plot2 + plot1 
combined_plot
#ggsave("baci_and_diff.jpg", height = 5, width = 9)
```

# fish meals metric
How many average size meals can the average size person eat in a month from uninvaded lakes?
-I first do it within each of the 4 invasion groupings
-I then do it by discounting zm post by the difference in uninvaded pre and post
```{r}
############## first discount zm after by the change in uninvaded ############
#within a draw find the difference pre and post invaded 
#discount
marg_draw_plot <- marg_draw_plot %>% 
  #hg_ppm (ug/g) * grams of meal = ug of mercury ingested
  mutate(hg_intake_mircogram = hg_ppm * 227) %>% 
  #safe ingestion of 0.1 ug/kg/day for a 80 kg individual - extrapolated to longer time periods 
  # ug/(kg*d) * kg = safe intake ug/time  
  mutate(max_intake_day = 0.1 * 80,
        max_intake_week = max_intake_day*7,
        max_intake_month = max_intake_day*30) %>% 
  #max meals per week = safe/ ug in meal
  mutate(safe_consumption_day = max_intake_day/hg_intake_mircogram,
         safe_consumption_week = max_intake_week/hg_intake_mircogram,
         safe_consumption_month = max_intake_month/hg_intake_mircogram)

discount <- marg_draw_plot %>% 
  select(species, interaction, .draw, hg_ppm) %>% 
    pivot_wider(id_cols = c(species, .draw),
                values_from = hg_ppm,
                names_from = interaction) %>% 
    mutate(ref_discount = `Uninvaded After` - `Uninvaded Before`,
           `Zebra Mussel After` = `Zebra Mussel After` - ref_discount) %>% 
    select(species, 
           .draw, 
           `Zebra Mussel Before`, 
           `Zebra Mussel After`) %>% 
    pivot_longer(cols = c(`Zebra Mussel After`, `Zebra Mussel Before`), 
                 names_to = c("interaction"),
                 values_to = c("hg_ppm"),
                ) %>% 
  #hg_ppm (ug/g) * grams of meal = ug of mercury ingested
  mutate(hg_intake_mircogram = hg_ppm * 227) %>% 
  #safe ingestion of 0.1 ug/kg/day for a 80 kg individual - extrapolated to longer time periods 
  # ug/(kg*d) * kg = safe intake ug/time  
  mutate(max_intake_day = 0.1 * 80,
        max_intake_week = max_intake_day*7,
        max_intake_month = max_intake_day*30) %>% 
  #max meals per week = safe/ ug in meal
  mutate(safe_consumption_day = max_intake_day/hg_intake_mircogram,
         safe_consumption_week = max_intake_week/hg_intake_mircogram,
         safe_consumption_month = max_intake_month/hg_intake_mircogram) %>% 
  mutate(interaction = as.factor(case_when(interaction == "Zebra Mussel Before" ~ "Before",
                                 TRUE ~ "After")),
         interaction = factor(interaction, levels = c("Before", "After"))) %>% 
  mutate(species = as.factor(case_when(species == "wae" ~ "Walleye",
                             species == "nop" ~ "Northern Pike",
                             species == "lmb" ~ "Largemouth Bass",
                             species == "blc" ~ "Black Crappie",
                             species == "yep" ~ "Yellow Perch",
                             species == "blg" ~ "Bluegill")), 
         species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")))

#meals summary
sum_meals <- discount %>% 
  group_by(species, interaction) %>% 
  summarise(median_meals_month = median(safe_consumption_month)) %>% 
  print(n = nrow(.))

#diff meals summary
sum_meals %>% 
  pivot_wider(id_cols = species,
              names_from = interaction,
              values_from = median_meals_month) %>% 
  mutate(diff_zm = After - Before) %>% 
  select(species, diff_zm)

#meals per week
discount %>% 
  ggplot(aes(interaction, safe_consumption_week, color = interaction)) +
  stat_pointinterval(.width = c(.66, .95)) +
  theme_minimal() +
  ylab("Safe Meals Per Week") +
  xlab("") +
  scale_color_manual(values = c("lightblue", "salmon")) +
  facet_wrap(~species, scales = "free") +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.title.y = element_text(size = 10),
    strip.background = element_rect(color = "black", fill = NA, linewidth  = 0.5),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5) 
  ) 

#meals per month
meals1 <- discount %>% 
  ggplot(aes(interaction, safe_consumption_month, color = interaction)) +
  stat_pointinterval(.width = c(.66, .95)) +
  theme_minimal() +
  ylab("Safe Meals Per Month") +
  xlab("") +
  scale_color_manual(values = c("lightblue", "salmon")) +
  facet_wrap(~species, scales = "free") +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.title.y = element_text(size = 10),
    strip.background = element_rect(color = "black", fill = NA, linewidth  = 0.5),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5) 
  ) 
meals1

species_images <- data.frame(
  species = c("Walleye", 
              "Northern Pike", 
              "Largemouth Bass", 
              "Black Crappie",
              "Yellow Perch", 
              "Bluegill"),
  img = c(file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "walleye_outline.jpg"), 
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "pike_outline.jpg"), 
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "largemouth_bass_outline.jpg"), 
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "crappie_outline.jpg"),
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "perch_outline.jpg"),
          file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "bluegill_outline.jpg")),
  x = 1.5,
  y = c(6.5, 4.25, 9, 18, 40, 27)
)

meals_icons <- meals1 + 
  geom_image(data = species_images, aes(y = y, x = x, image = img), size = .3, inherit.aes = F)
meals_icons
ggsave("diff_meals.jpg", height = 5, width = 7, units = "in", dpi = 600)

rm(marg_draw_plot, sum_meals, marginal_effects_all, discount)

###############code below to make for each level of the baci#################
#lets use those marginal effects to calculate this - and keep track of our units as we go
marg_draw_plot <- marg_draw_plot %>% 
  #hg_ppm (ug/g) * grams of meal = ug of mercury ingested
  mutate(hg_intake_mircogram = hg_ppm * 227) %>% 
  #safe ingestion of 0.1 ug/kg/day for a 80 kg individual - extrapolated to longer time periods 
  # ug/(kg*d) * kg = safe intake ug/time  
  mutate(max_intake_day = 0.1 * 75,
        max_intake_week = max_intake_day*7,
        max_intake_month = max_intake_day*30) %>% 
  #max meals per week = safe/ ug in meal
  mutate(safe_consumption_day = max_intake_day/hg_intake_mircogram,
         safe_consumption_week = max_intake_week/hg_intake_mircogram,
         safe_consumption_month = max_intake_month/hg_intake_mircogram)
glimpse(marg_draw_plot)

#meals summary
sum_meals <- marg_draw_plot %>% 
  group_by(species, interaction) %>% 
  summarise(median_meals_month = median(safe_consumption_month)) %>% 
  print(n = nrow(.))

#diff meals summary
sum_meals %>% 
  pivot_wider(id_cols = species,
              names_from = interaction,
              values_from = median_meals_month) %>% 
  mutate(diff_uninvaded = `Uninvaded After` - `Uninvaded Before`,
         diff_zm = `Zebra Mussel After` - `Zebra Mussel Before`) %>% 
  select(species, diff_uninvaded, diff_zm)

#plots 
marg_draw_plot %>% 
  ggplot(aes(interaction, safe_consumption_week, color = interaction)) +
  stat_pointinterval(.width = c(.66, .95)) +
  scale_color_manual("Lake Invasion", 
                      values = c("lightblue", "lightblue", "salmon", "salmon")) +
  theme_minimal() +
  ylab("Safe Meals Per Week") +
  xlab("") +
  facet_wrap(~species, scales = "free") +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.title.y = element_text(size = 7)
  ) 
```

# Map 
- I first make a map of any lake included in the study, regardless of species 
- I then make species specific maps
- I tried to combine them but it will probably be easier to do that outside of R or leave them separate 
```{r}
lake_locs <- combined_data %>% 
  distinct(dow, waterway, centroid_lat, centroid_lon, zm_lake)

#make sf objects
lakes_sf <- st_as_sf(lake_locs, coords = c("centroid_lon", "centroid_lat"), crs = 4326)

mn_map <- states(cb = T) %>% 
  filter(STUSPS == "MN")

minnesota_sf <- st_as_sf(mn_map)
rm(mn_map)


#simple map showing lake locations by species 
mn_map <- ggplot() +
  # Plot Minnesota state
  geom_sf(data = minnesota_sf, fill = "white", color = "black") +
  
  # Plot lakes with centroids
  geom_sf(data = lakes_sf, 
          aes(color = zm_lake),
          size = 3) +
  scale_color_manual("Invasion Status of Lake", values = c("lightblue", "salmon"), labels = c("Uninvaded", "Zebra Mussel")) +
  # Remove gridlines and set white background
  theme_minimal() +
  theme(panel.background = element_rect(fill = "white", color = NA),
        panel.grid = element_blank(),
        legend.position = c(.79, .4),
         legend.box.background = element_rect(color = "black", size = 0.5),
        legend.key.size = unit(.2, "cm")) +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "br", 
                         which_north = "true",
                         pad_x = unit(0.55, "cm"),
                         pad_y = unit(0.7, "cm"))
mn_map
ggsave("BACI_lakes_map.jpg", height = 7, width = 7, dpi = 600, bg = "white")

#with inset
us_sf <- st_as_sf(maps::map("state", plot = F, fill = T))

us_map <- ggplot() +
  geom_sf(data = us_sf, fill = "gray90", color = "black", size = 0.3) +
  geom_sf(data = minnesota_sf, fill = "red", color = "black") +
  theme_void()

inset_x <- 0.7
inset_y <- 0.8
inset_width <- 0.22
inset_height <- 0.15

mn_map_total <- ggdraw(mn_map) +
  draw_plot(us_map, x = 0.7, y = 0.76, width = 0.22, height = 0.22) +
  draw_line(x = c(inset_x, inset_x + inset_width), y = c(inset_y, inset_y), color = "black", size = 1) +  
  draw_line(x = c(inset_x, inset_x + inset_width), y = c(inset_y + inset_height, inset_y + inset_height), color = "black", size = 1) +  
  draw_line(x = c(inset_x, inset_x), y = c(inset_y, inset_y + inset_height), color = "black", size = 1) +  
  draw_line(x = c(inset_x + inset_width, inset_x + inset_width), y = c(inset_y, inset_y + inset_height), color = "black", size = 1)  
mn_map_total
ggsave("BACI_lakes_map_inset.jpg", height = 7, width = 7, dpi = 600, bg = "white")

########################species specific maps############################
lake_locs_spec <- combined_data %>% 
  distinct(spec, dow, waterway, centroid_lat, centroid_lon, zm_lake) %>% 
  mutate(species = as.factor(case_when(spec == "NOP" ~ "Northern Pike",
                             spec == "WAE" ~ "Walleye",
                             spec == "BLG" ~ "Bluegill",
                             spec == "BLC" ~ "Black Crappie",
                             spec == "YEP" ~ "Yellow Perch",
                             spec == "LMB" ~ "Largemouth Bass",
                             TRUE ~ spec)),
         species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")))

#make sf objects
lakes_sf <- st_as_sf(lake_locs_spec, coords = c("centroid_lon", "centroid_lat"), crs = 4326)

mn_map <- states(cb = T) %>% 
  filter(STUSPS == "MN")

minnesota_sf <- st_as_sf(mn_map)
rm(mn_map)

species_images <- lake_locs_spec %>%
  group_by(species) %>%
  summarise(y = max(centroid_lat, na.rm = TRUE) * 1.02,  # Slightly above the highest lake
            x = max(centroid_lon, na.rm = TRUE) * 1.02) %>%  # Slightly to the right
  mutate(img = case_when(
    species == "Walleye" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "walleye_outline.jpg"),
    species == "Northern Pike" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "pike_outline.jpg"),
    species == "Largemouth Bass" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "largemouth_bass_outline.jpg"),
    species == "Black Crappie" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "crappie_outline.jpg"),
    species == "Yellow Perch" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "perch_outline.jpg"),
    species == "Bluegill" ~ file.path("D:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", "Statewide mercury modeling", "Zebra Mussel Baci Analysis", "Data", "bluegill_outline.jpg")
  ),
  y = 49,
  x = -91)  


#simple map showing lake locations by species 
mn_map_spec <- ggplot() +
  # Plot Minnesota state
  geom_sf(data = minnesota_sf, fill = "white", color = "black") +
  
  # Plot lakes with centroids
  geom_sf(data = lakes_sf, 
          aes(color = zm_lake),
          size = 2) +
  scale_color_manual("Invasion Status of Lake", values = c("lightblue", "salmon"), labels = c("Uninvaded", "Zebra Mussel")) +
  # Remove gridlines and set white background
  theme_minimal() +
  theme(panel.background = element_rect(fill = "white", color = NA),
        panel.grid = element_blank(),
        #legend.position = c(.79, .4),
        legend.position = "none",
         legend.box.background = element_rect(color = "black", size = 0.5),
        legend.key.size = unit(.2, "cm"),
        strip.background = element_rect(color = "black", fill = NA, linewidth  = 0.5),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5) 
        ) +
  # annotation_scale(location = "br") +
  # annotation_north_arrow(location = "br", 
  #                        which_north = "true",
  #                        pad_x = unit(0.55, "cm"),
  #                        pad_y = unit(0.7, "cm")) +
  facet_wrap(~ species)
mn_map_spec

mn_map_icons <- mn_map_spec +
  geom_image(data = species_images, aes(x = x, y = y, image = img), size = 0.3) +
  ylab("") +
  xlab("")
mn_map_icons
ggsave("map_spec_icons.jpg", width = 7, height = 5, dpi = 600)

#########################combined maps#######################################
combined_map <- mn_map_total / mn_map_icons+
  plot_layout(heights = c(1.7, 1))
combined_map
ggsave("combined_maps.jpg", width = 5, height = 7, dpi = 600)

#############remove objects######################
rm(lake_summary, lakes_sf, minnesota_sf, spec, inset_height, inset_width, inset_x, inset_y, mn_map, us_sf, us_map)
```

# Probability of exceedance & length vs hgppm 
```{r}
length_ranges <- combined_data %>%
  group_by(spec) %>% 
  summarize(min_length = min(scaled_length),
            max_length = max(scaled_length))

length_seq <- length_ranges %>% 
  rowwise() %>% 
  mutate(scaled_length = list(seq(min_length, max_length, length.out = 200))) %>% 
  unnest(cols = c(scaled_length))
rm(length_ranges)

length.stats <- combined_data %>%
  group_by(spec) %>% 
  summarise(mean.length=mean(lgthmm), sd.length=sd(lgthmm), min.length=min(lgthmm), max.length=max(lgthmm))

pred_data <- expand_grid(spec = unique(length_seq$spec),
                         zm_lake = c("y", "n"),
                         zm_sample_lag_2 = c("y", "n")) %>% 
  left_join(length_seq, by = "spec") %>% 
  ungroup()
    
get_prob_exceed_spec <- function(spec_filter, fit_model, pred_data, length_stats) {
  pred_data_spec <- pred_data %>% 
    filter(spec == spec_filter) %>% 
    filter(scaled_length <= 3 & scaled_length >= -3)
  
  length_stats_spec <- length_stats %>% 
    filter(spec == spec_filter)
  
  print(paste("Glancing at filtered pred_data_spec for species:", spec_filter))
  glimpse(pred_data_spec) 
  
  print(paste("Glancing at filtered length_stats_spec for species:", spec_filter))
  glimpse(length_stats_spec) 
  
  print(paste("checking selected model fit:", spec_filter))
  print(summary(fit_model))
  
  mean_length <- length_stats_spec$mean.length
  sd_length <- length_stats_spec$sd.length
  
  prob_exceed <- fit_model %>% 
    add_predicted_draws(newdata = pred_data_spec, re_formula = NA) %>% 
    mutate(invasion = paste0(zm_lake, zm_sample_lag_2)) %>% 
    mutate(mean_length = mean_length,
           sd_length = sd_length,
           lgthmm = (scaled_length * sd_length) + mean_length,
           is_above_threshold = .prediction > log(0.22)) %>% 
    group_by(invasion, lgthmm) %>% 
    summarise(prob = mean(is_above_threshold), .groups = "drop") %>% 
    mutate(spec = spec_filter) %>% 
    pivot_wider(id_cols = lgthmm, names_from = invasion, values_from = prob) %>% 
    mutate(yy_adjusted = yy - (ny - nn)) %>% 
    pivot_longer(cols = c(yy, ny, nn, yn, yy_adjusted), names_to = "invasion", values_to = "prob") %>% 
    mutate(species = spec_filter)
  return(prob_exceed)
}

#walleye
wae_exceed <- get_prob_exceed_spec(spec_filter = "WAE", 
                                   fit_model = fit_wae, 
                                   pred_data = pred_data,
                                   length_stats = length.stats)
#pike
nop_exceed <- get_prob_exceed_spec(spec = "NOP", 
                                   fit_model = fit_nop, 
                                   pred_data = pred_data,
                                   length_stats = length.stats)
#bluegill
blg_exceed <- get_prob_exceed_spec(spec = "BLG", 
                                   fit_model = fit_blg, 
                                   pred_data = pred_data,
                                   length_stats = length.stats)
#crappie
blc_exceed <- get_prob_exceed_spec(spec = "BLC", 
                                   fit_model = fit_blc, 
                                   pred_data = pred_data,
                                   length_stats = length.stats)
#largemouth
lmb_exceed <- get_prob_exceed_spec(spec = "LMB", 
                                   fit_model = fit_lmb, 
                                   pred_data = pred_data,
                                   length_stats = length.stats)

#perch
yep_exceed <- get_prob_exceed_spec(spec = "YEP", 
                                   fit_model = fit_yep, 
                                   pred_data = pred_data,
                                   length_stats = length.stats)

combined_exceed <- bind_rows(wae_exceed, 
                             nop_exceed,
                             yep_exceed,
                             blg_exceed,
                             blc_exceed,
                             lmb_exceed)

exceed_plot <- combined_exceed %>% 
  filter(invasion %in% c("yy_adjusted", "yn")) %>% 
  mutate(invasion = case_when(invasion == "yy_adjusted" ~ "yy",
                              TRUE ~ invasion),
         species = as.factor(case_when(species == "BLC" ~ "Black Crappie",
                                       species == "BLG" ~ "Bluegill",
                                       species == "LMB" ~ "Largemouth Bass",
                                       species == "NOP" ~ "Northern Pike",
                                       species == "WAE" ~ "Walleye",
                                       species == "YEP" ~ "Yellow Perch",
                                       TRUE ~ NA)),
         species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")),
         prob = round(prob, 2))

vert_line <- exceed_plot %>% 
  group_by(species) %>% 
  mutate(distance = case_when(species == "Walleye" ~ abs(lgthmm - 381),
                              species == "Northern Pike" ~ abs(lgthmm - 610),
                              species == "Largemouth Bass" ~ abs(lgthmm - 305),
                              species == "Black Crappie" ~ abs(lgthmm - 205),
                              species == "Yellow Perch" ~ abs(lgthmm - 205),
                              species == "Bluegill" ~ abs(lgthmm - 155))) %>% 
  filter(distance == min(distance)) %>% 
  ungroup()

prob_plot <- ggplot() +
  geom_smooth(data = exceed_plot, aes(x = lgthmm, y = prob, color = invasion), method = "loess", span = .3, se = F) + 
  geom_point(data = vert_line, aes(x =lgthmm, y = prob, color = invasion)) +
  geom_line(data = vert_line, aes(x = lgthmm, y = prob #, group = invasion
                                  )) +
  facet_wrap(~species, scales = "free_x") +
  scale_colour_manual(values=c("#00CCCC", "#FF6666"), labels=c("Before", "After")) +
  ylim(c(-0.01,1.01)) +
  xlab("Length (mm)") +
  ylab("Probability of Exceeding 0.22 ppm") +
  theme_bw() +
  theme(legend.position = c(.25, .65), 
        legend.title = element_blank(), 
        legend.background = element_blank(), 
        strip.background = element_rect(color = "black", fill = NA, linewidth  = 0.5),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
        panel.grid = element_blank())
prob_plot

species_images <- data.frame(
  species = as.factor(c("Northern Pike", 
                        "Walleye",
              "Largemouth Bass", 
              "Black Crappie",
              "Yellow Perch", 
              "Bluegill")),
  img = c(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "pike_outline.jpg"), 
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "walleye_outline.jpg"), 
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "largemouth_bass_outline.jpg"), 
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "crappie_outline.jpg"),
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "perch_outline.jpg"),
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "bluegill_outline.jpg")),
  x = c(350, 250, 250, 170, 155, 155),
  y = 0.8) 

prob_plot_icons <- prob_plot +
  geom_image(data = species_images, aes(y = y, x = x, image = img), size = .3, inherit.aes = F)
prob_plot_icons
ggsave("prob_exceed.jpg", height = 5, width = 7, dpi = 600)

rm(nop_exceed,
   wae_exceed,
   lmb_exceed,
   blc_exceed,
   yep_exceed,
   blg_exceed,
   combined_exceed,
   vert_line,
   prob_plot,
   prob_plot_icons,
   species_images,
   exceed_plot)

########################ppm vs length###################################
get_expected_hgppm <- function(spec_filter, fit_model, pred_data, length_stats) {
  pred_data_spec <- pred_data %>% 
    filter(spec == spec_filter) %>% 
    filter(scaled_length >= -5 & scaled_length <= 5)
  
  length_stats_spec <- length_stats %>% 
    filter(spec == spec_filter)
  
  print(paste("Glancing at filtered pred_data_spec for species:", spec_filter))
  glimpse(pred_data_spec) 
  
  print(paste("Glancing at filtered length_stats_spec for species:", spec_filter))
  glimpse(length_stats_spec) 
  
  print(paste("checking selected model fit:", spec_filter))
  print(summary(fit_model))
  
  mean_length <- length_stats_spec$mean.length
  sd_length <- length_stats_spec$sd.length
  
  expected_hgppm <- fit_model %>% 
    add_predicted_draws(newdata = pred_data_spec, re_formula = NA) %>% 
    mutate(mean_length = mean_length,
           sd_length = sd_length,
           lgthmm = (scaled_length * sd_length) + mean_length) %>% 
    mutate(invasion = paste0(zm_lake, zm_sample_lag_2)) %>% 
    group_by(invasion, lgthmm) %>% 
    summarise(median=median(exp(.prediction)), 
              lci=quantile(exp(.prediction), .025), 
              uci=quantile(exp(.prediction), 0.975)) %>% 
    ungroup() %>% 
    pivot_wider(id_cols = lgthmm, names_from = invasion, values_from = c(median, lci, uci)) %>% 
    mutate(median_correction = median_ny - median_nn,
           median_post_corrected = median_yy - median_correction,
           uci_correction = uci_ny - uci_nn,
           uci_post_corrected = uci_yy - uci_correction,
           lci_correction = lci_ny - lci_nn,
           lci_post_corrected = lci_yy - lci_correction) %>% 
    pivot_longer(cols = -lgthmm, 
                 names_to = c("name", "invasion"),
                 names_sep = "_",
                 values_to = "value") %>%
    pivot_wider(id_cols = c(lgthmm, invasion),
                names_from = name,
                values_from = value) %>%
    group_by(invasion) %>% 
    mutate(smoothed_uci = predict(loess(uci ~ lgthmm, span = 0.5)),
         smoothed_lci = predict(loess(lci ~ lgthmm, span = 0.5)),
         smoothed_median = predict(loess(median ~ lgthmm, span = 0.5))) %>%
    ungroup() %>% 
    mutate(species = spec_filter) %>% 
    filter(invasion %in% c("post", "yn")) %>% 
    mutate(invasion = case_when(invasion == "post" ~ "yy",
                                TRUE ~ invasion))
  return(expected_hgppm)
}

nop_expect <- get_expected_hgppm(spec_filter = "NOP",
                                 fit_model = fit_nop, 
                                 pred_data = pred_data,
                                 length_stats = length.stats)

wae_expect <- get_expected_hgppm(spec_filter = "WAE",
                                 fit_model = fit_wae, 
                                 pred_data = pred_data,
                                 length_stats = length.stats)

lmb_expect <- get_expected_hgppm(spec_filter = "LMB",
                                 fit_model = fit_lmb, 
                                 pred_data = pred_data,
                                 length_stats = length.stats)

blc_expect <- get_expected_hgppm(spec_filter = "BLC",
                                 fit_model = fit_blc, 
                                 pred_data = pred_data,
                                 length_stats = length.stats)
yep_expect <- get_expected_hgppm(spec_filter = "YEP",
                                 fit_model = fit_yep, 
                                 pred_data = pred_data,
                                 length_stats = length.stats)
blg_expect <- get_expected_hgppm(spec_filter = "BLG",
                                 fit_model = fit_blg, 
                                 pred_data = pred_data,
                                 length_stats = length.stats)

combined_expect <- bind_rows(nop_expect,
                             wae_expect,
                             lmb_expect,
                             blc_expect,
                             yep_expect,
                             blg_expect) %>% 
  mutate(species = as.factor(case_when(species == "BLC" ~ "Black Crappie",
                                       species == "BLG" ~ "Bluegill",
                                       species == "LMB" ~ "Largemouth Bass",
                                       species == "NOP" ~ "Northern Pike",
                                       species == "WAE" ~ "Walleye",
                                       species == "YEP" ~ "Yellow Perch",
                                       TRUE ~ NA)),
         species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill")))

combined_data %>% 
  group_by(spec) %>% 
  summarise(max(lgthmm))

#length vs ppm plot
hgppm_plot <- ggplot() +
   geom_ribbon(data = combined_expect, aes(x = lgthmm, y = median, color = invasion, ymax = smoothed_uci, ymin = smoothed_lci, fill = invasion), lwd = 0, alpha = .2) +
   geom_line(data = combined_expect, lwd = 1, aes(x = lgthmm, y = median, color = invasion, lty = invasion)) +
  geom_point(data = combined_data %>% 
               filter(zm_lake == "y") %>% 
               mutate(invasion = case_when(zm_sample_lag_2 == "y" ~ "yy",
                                           zm_sample_lag_2 == "n" ~ "yn"),
                      
                      species = as.factor(case_when(spec == "BLC" ~ "Black Crappie",
                                       spec == "BLG" ~ "Bluegill",
                                       spec == "LMB" ~ "Largemouth Bass",
                                       spec == "NOP" ~ "Northern Pike",
                                       spec == "WAE" ~ "Walleye",
                                       spec == "YEP" ~ "Yellow Perch",
                                       TRUE ~ NA)),
                      species = factor(species, levels = c("Northern Pike",
                                              "Walleye",
                                              "Largemouth Bass",
                                              "Black Crappie",
                                              "Yellow Perch",
                                              "Bluegill"))), aes(x = lgthmm, y = hgppm,
                                                                               color = invasion),
                      alpha = .2) +
               facet_wrap(~species, scales = "free") +
  scale_linetype_manual("Impact", values = c(1, 6), labels = c("Before", "After")) + 
  scale_colour_manual("Impact", values = c("#00CCCC", "#FF6666"), labels = c("Before", "After")) +
  scale_fill_manual("Impact", values = c("#00CCCC", "#FF6666"), labels = c("Before", "After")) +
  ylab("Mercury Concentration (ppm)") +
  xlab("Fish Length (mm)") +
  geom_hline(yintercept = 0.22,linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = c(.06, .82), 
        legend.title = element_blank(), 
        legend.background = element_blank(), 
        strip.background = element_rect(color = "black", fill = NA, linewidth  = 0.5),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5), 
        panel.grid = element_blank())
hgppm_plot

species_images <- data.frame(
  species = as.factor(c("Northern Pike", 
                        "Walleye",
              "Largemouth Bass", 
              "Black Crappie",
              "Yellow Perch", 
              "Bluegill")),
  img = c(file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "pike_outline.jpg"), 
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "walleye_outline.jpg"), 
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "largemouth_bass_outline.jpg"), 
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "crappie_outline.jpg"),
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "perch_outline.jpg"),
          file.path("G:", "Shared drives", "Hansen Lab", "RESEARCH PROJECTS", 
                    "Statewide mercury modeling", "Zebra Mussel Baci Analysis", 
                    "Data", "bluegill_outline.jpg")),
  x = c(355, 255, 250, 168, 155, 150),
  y = c(2.2, 1.8, 0.75, 0.5, 0.3, 0.55)) 

hgppm_plot_icons <- hgppm_plot +
  geom_image(data = species_images, aes(y = y, x = x, image = img), size = .3, inherit.aes = F)
hgppm_plot_icons
ggsave("hgppm_raw_predicted_lines.jpg", width = 7, height = 5)


#stat for when fish cross the line?
combined_expect %>% 
  mutate(distance = abs(median - 0.22)) %>% 
  group_by(invasion, species) %>% 
  filter(distance == min(distance)) %>%
  select(species, lgthmm, invasion) %>% 
  pivot_wider(names_from = c(invasion),
              values_from = lgthmm) %>% 
  mutate(percent_change_zm = ((yy-yn)/yn)*100,
         diff = yy-yn) %>% 
  write_csv("percent_change_crossing_lengths.csv")

rm(pred_data,
   nop_expect,
   wae_expect,
   lmb_expect,
   blc_expect,
   blg_expect,
   yep_expect,
   hgppm_plot_icons,
   hgppm_plot,
   combined_expect,
   species_images)
```

# Data Coverage Stats
```{r}
#stats for the singular model run of 2 lag years ref and zm with matching
lakes <- combined_data %>% 
  group_by(spec, zm_lake, zm_sample_lag_2) %>% 
  summarise(n_lakes = n_distinct(dow),
            n_years = n_distinct(yearcoll),
            n_samples = n()) %>% 
  print(n = nrow(.))
write_csv(lakes, "baci_lake_stats.csv")
rm(lakes)


fish <- combined_data %>% 
  group_by(spec, zm_lake, zm_sample_lag_2) %>% 
  summarise(mean = mean(lgthmm),
            sd = sd(lgthmm),
            min = min(lgthmm),
            max = max(lgthmm),
            n = n()) %>% 
  print(n = nrow(.))
write_csv(fish, "baci_fish_pre_post_stats.csv")
rm(fish)

fish <- combined_data %>% 
  group_by(spec, zm_lake) %>% 
  summarise(mean = mean(lgthmm),
            sd = sd(lgthmm),
            min = min(lgthmm),
            max = max(lgthmm),
            n = n())
write_csv(fish, "baci_fish_invasion_stats.csv")
rm(fish)

fish <- combined_data %>% 
  group_by(spec) %>% 
  summarise(mean = mean(lgthmm),
            sd = sd(lgthmm),
            min = min(lgthmm),
            max = max(lgthmm),
            n_lakes = n_distinct(dow),
            n = n())
write_csv(fish, "baci_fish_spec_stats.csv")
rm(fish)

#limit of detection?
combined_data %>% 
  filter(hgcode == "K") %>% 
  group_by(hgcode, hgppm, spec) %>% 
  count() %>% 
  print(n = nrow(.))

#how many unique lakes across species
combined_data %>% 
  group_by(zm_lake) %>% 
  summarise(n_distinct(dow))

combined_data %>% 
  group_by(spec, zm_lake) %>% 
  summarise(n_distinct(dow))

```

# Model summaries 
```{r}

```


################below need to be updated to the data################

# Elapsed time summaries
```{r}
# Define species list
species_list <- c("WAE", "NOP", "BLG", "BLC", "LMB", "YEP", "WTS")

# Create an empty list to store plots
plots <- list()

# Loop over species
for (species in species_list) {
  
  # Step 1: Create species-specific data
  species_data <- create_species_data(species = species, data = hg.model, lag_variable = "zm_sample_lag_2")
  
  # Step 2: Process infestation year and post_inv flag
  species_data <- species_data %>%
    mutate(year_infested = ifelse(is.na(year_infested), nearest_infested_year, year_infested)) %>%
    mutate(year_inf_lag2 = year_infested + 2) %>% # time lag set to 2 years
    mutate(post_inv = yearcoll > year_inf_lag2)
  
  # Step 3: Calculate pre/post mean year values
  lake_deltas <- species_data %>%
    group_by(waterway, dow, post_inv, zm_lake, year_inf_lag2) %>%
    summarize(
      mean_yr = mean(yearcoll),
      n_yr = length(unique(yearcoll)),
      .groups = "drop"
    ) %>%
    pivot_wider(
      names_from = post_inv, 
      values_from = c(mean_yr, n_yr),
      names_glue = "{.value}_post_inv_{post_inv}"
    ) %>%
    mutate(delta_t = round(mean_yr_post_inv_TRUE - mean_yr_post_inv_FALSE, 2))
  
  # Step 4: Compute max count for scaling density plot
  max_count <- max(ggplot_build(ggplot(lake_deltas, aes(x = delta_t)) + 
                                  geom_histogram(binwidth = 1))$data[[1]]$count, na.rm = TRUE)
  
  # Step 5: Create the plot
  p <- ggplot(lake_deltas, aes(x = delta_t, fill = zm_lake)) +
    geom_histogram(aes(y = ..count..), binwidth = 1, alpha = 0.4, position = "identity", show.legend = FALSE) +
    geom_density(aes(y = ..density.. * 10* max_count, color = zm_lake), alpha = 0.01, size = 1, show.legend = species == "WTS") +  # Only show legend for the first plot
    scale_y_continuous(
      name = "Histogram Count",
      sec.axis = sec_axis(~ . / max_count, name = "Density x10")
    ) +
    labs(x = "Delta T", fill = "ZM Lake", color = "ZM Lake", title = species) +
    scale_color_manual("", values = c("lightblue", "salmon"), labels = c("Uninvaded", "Zebra Mussel")) +
    scale_fill_manual("", values = c("lightblue", "salmon"), labels = c("Uninvaded", "Zebra Mussel")) +
    theme_minimal() +
    theme(#legend.position = ifelse(species == "WTS", "right", "none"),
          legend.position = "none"
          )  # Only show legend for the first plot

  
  # Store the plot in a list
  plots[[species]] <- p
  

}

# Now `plots` contains a separate ggplot object for each species
# Arrange the plots in a 2-column, 3-row grid (adjusted for 6 species)
wrap_plots(plots)
ggsave("delta_t_raw.jpg", width = 11, height = 7)

#########per year marginal effect - lets just start with walleye###########
fit_wae <- readRDS("Output/BACI/baci_update_wae.rds")

group_deltas <- wae_data %>% 
  group_by(zm_lake, post_inv) %>% 
  summarize(
    mean_yr = mean(yearcoll), 
    n_yr = length(unique(yearcoll)), 
    .groups = "drop"
  ) %>% 
  pivot_wider(
    names_from = post_inv, 
    values_from = c(mean_yr, n_yr),
    names_glue = "{.value}_post_inv_{post_inv}"
  ) %>% 
  mutate(delta_t = round(mean_yr_post_inv_TRUE - mean_yr_post_inv_FALSE, 2)) %>% 
  mutate(zm_lake = case_when(zm_lake == "n" ~ "Uninvaded",
                             zm_lake == "y" ~ "Zebra Mussel"))

#marginal effects based on specific length
length_stats <-  wae_data %>% 
  summarise(mean_length = mean(lgthmm),
            sd_length = sd(lgthmm))
specific_length_scaled <- (381 - length_stats$mean_length) / length_stats$sd_length

marg_draws <- emmeans(fit_wae, ~ zm_lake * zm_sample_lag_2, 
                      at = list(scaled_length = specific_length_scaled), 
                      re_fromula = NA, 
                      epred = T) %>% 
  gather_emmeans_draws() %>% 
  mutate(hg_ppm = exp(.value))


#creating differences
contrasts_draws <- marg_draws %>% 
  select(-.chain, -.iteration) %>% 
  pivot_wider(names_from = c(zm_lake, zm_sample_lag_2),
              values_from = c(.value, hg_ppm))

diff_lakes <- contrasts_draws %>% 
  mutate(diff_reference = hg_ppm_n_y - hg_ppm_n_n,
         diff_zm = hg_ppm_y_y - hg_ppm_y_n) %>% 
  select(.draw, diff_reference, diff_zm) %>% 
  pivot_longer(cols = c(diff_reference, diff_zm),
               names_to = "zm_lake",
               values_to = "value") %>% 
  mutate(zm_lake = case_when(zm_lake == "diff_zm" ~ "Zebra Mussel",
                             TRUE ~ "Uninvaded"))


diff_lakes <- diff_lakes %>% 
  left_join(group_deltas, by = "zm_lake") %>% 
  mutate(value_per_year = value/delta_t)

diff_of_diff <- diff_lakes %>% 
  pivot_wider(names_from = zm_lake,
              id_cols = .draw,
              values_from = c("value", "value_per_year")) %>% 
  mutate(diff = `value_Zebra Mussel` - value_Uninvaded,
         diff_delta = `value_per_year_Zebra Mussel` - value_per_year_Uninvaded) %>% 
  select(.draw, diff, diff_delta) %>% 
  pivot_longer(cols = c(diff, diff_delta),
               names_to = "type",
               values_to = "values")


diff_of_diff %>% 
  ggplot() +
  stat_pointinterval(aes(x = type, y = values, color = type), .width = c(.66, .95)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  ylab("Difference in Mercury Concentration\nBefore and After (ppm)") +
  xlab("") +
  scale_color_manual(values = c("lightblue", "salmon", "black")) +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        #axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.title.y = element_text(size = 8)) 


species_list <- c("wae", "nop", "blg", "blc", "lmb", "yep", "wts")

process_species <- function(species) {
  # Load model
  fit_species <- readRDS(paste0("Output/BACI/baci_update_", species, ".rds"))
  
  # Load species data
  species_data <- get(paste0(species, "_data")) # Assuming each dataset is named accordingly
  
  species_data <- species_data %>%
    mutate(year_infested = ifelse(is.na(year_infested), nearest_infested_year, year_infested)) %>%
    mutate(year_inf_lag2 = year_infested + 2) %>% # time lag set to 2 years
    mutate(post_inv = yearcoll > year_inf_lag2)
  
  group_deltas <- species_data %>% 
    group_by(zm_lake, post_inv) %>% 
    summarize(
      mean_yr = mean(yearcoll), 
      n_yr = length(unique(yearcoll)), 
      .groups = "drop"
    ) %>% 
    pivot_wider(
      names_from = post_inv, 
      values_from = c(mean_yr, n_yr),
      names_glue = "{.value}_post_inv_{post_inv}"
    ) %>% 
    mutate(delta_t = round(mean_yr_post_inv_TRUE - mean_yr_post_inv_FALSE, 2)) %>% 
    mutate(zm_lake = case_when(zm_lake == "n" ~ "Uninvaded",
                               zm_lake == "y" ~ "Zebra Mussel"))
  
  # Marginal effects based on specific length
  length_stats <- species_data %>% 
    summarise(mean_length = mean(lgthmm),
              sd_length = sd(lgthmm))
  specific_length_scaled <- (381 - length_stats$mean_length) / length_stats$sd_length
  
  marg_draws <- emmeans(fit_species, ~ zm_lake * zm_sample_lag_2, 
                        at = list(scaled_length = specific_length_scaled), 
                        re_formula = NA, 
                        epred = TRUE) %>% 
    gather_emmeans_draws() %>% 
    mutate(hg_ppm = exp(.value))
  
  # Creating differences
  contrasts_draws <- marg_draws %>% 
    select(-.chain, -.iteration) %>% 
    pivot_wider(names_from = c(zm_lake, zm_sample_lag_2),
                values_from = c(.value, hg_ppm))
  
  diff_lakes <- contrasts_draws %>% 
    mutate(diff_reference = hg_ppm_n_y - hg_ppm_n_n,
           diff_zm = hg_ppm_y_y - hg_ppm_y_n) %>% 
    select(.draw, diff_reference, diff_zm) %>% 
    pivot_longer(cols = c(diff_reference, diff_zm),
                 names_to = "zm_lake",
                 values_to = "value") %>% 
    mutate(zm_lake = case_when(zm_lake == "diff_zm" ~ "Zebra Mussel",
                               TRUE ~ "Uninvaded"))
  
  diff_lakes <- diff_lakes %>% 
    left_join(group_deltas, by = "zm_lake") %>% 
    mutate(value_per_year = value/delta_t)
  
  diff_of_diff <- diff_lakes %>% 
    pivot_wider(names_from = zm_lake,
                id_cols = .draw,
                values_from = c("value", "value_per_year")) %>% 
    mutate(diff = `value_Zebra Mussel` - value_Uninvaded,
           diff_delta = `value_per_year_Zebra Mussel` - value_per_year_Uninvaded) %>% 
    select(.draw, diff, diff_delta) %>% 
    pivot_longer(cols = c(diff, diff_delta),
                 names_to = "type",
                 values_to = "values")
  
  # Plot
  p <- diff_of_diff %>% 
    ggplot() +
    stat_pointinterval(aes(x = type, y = values, color = type), .width = c(.66, .95)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    ylab("Difference in Mercury Concentration\nBefore and After (ppm)") +
    xlab("") +
    scale_color_manual(values = c("lightblue", "salmon", "black")) +
    theme_minimal() +
    theme(legend.position = "none",
          panel.grid.major.x = element_blank(),
          axis.title.y = element_text(size = 8)) +
    ggtitle(species)
  
  return(p)
}

# Apply function over all species and store plots in a list
species_plots <- map(species_list, process_species)

wrap_plots(species_plots)
ggsave("delta_t_marg_effect.jpg", width = 11, height = 7)


#####time since invasion########################
all_species <- bind_rows(blc_data, 
                         blg_data,
                         lmb_data,
                         nop_data,
                         wae_data,
                         wts_data,
                         yep_data) %>% 
  mutate(impact_year = case_when(is.na(year_infested) ~ nearest_infested_year,
                                 TRUE ~ year_infested),
         time_since_invasion = yearcoll - impact_year,
         impact_yn = case_when(time_since_invasion >= 0 ~ "y",
                               TRUE ~ "n"),
         impact_yn_lag2 = case_when(time_since_invasion - 2 >= 0 ~ "y",
                                    TRUE ~ "n"))



#here we go - lake level 
lake_level_data <- all_species %>%
  mutate(impact_groups = paste0(zm_lake, zm_lake)) %>% 
  group_by(dow, spec, impact_groups) %>%
  summarise(time_since_invasion = mean(time_since_invasion), .groups = "drop")

max_count <- max(ggplot_build(
  ggplot(lake_level_data, aes(x = time_since_invasion)) + 
    geom_histogram(binwidth = 1)
  )$data[[1]]$count, na.rm = TRUE)

all_species %>% 
  mutate(impact_groups = paste0(zm_lake)) %>% 
  group_by(dow, spec, impact_groups) %>% 
  summarise(time_since_invasion = mean(time_since_invasion)) %>% 
  ggplot(aes(x = time_since_invasion, fill = impact_groups)) +
  geom_histogram(binwidth = 1, alpha = 0.4, position = "identity", show.legend = FALSE) +
  geom_density(aes(y = after_stat(density) * max_count * 1, color = impact_groups), # Scaling density
               alpha = 0.01, linewidth = 1, show.legend = FALSE) +
  scale_y_continuous(
    name = "Histogram Count",
    sec.axis = sec_axis(~ . / max_count, name = "Density (scaled)")
  ) +
  geom_vline(xintercept = 0) +
  scale_color_manual("", values = c("lightblue", "salmon"), labels = c("Uninvaded", "Zebra Mussel")) +
  scale_fill_manual("", values = c("lightblue", "salmon"), labels = c("Uninvaded", "Zebra Mussel")) +
  theme_minimal() +
  facet_wrap(~spec, scales = "free")

#now do it pre and post for each lake
lake_level_data <- all_species %>%
  mutate(impact_groups = paste0(zm_lake, impact_yn)) %>% 
  group_by(dow, spec, impact_groups) %>%
  summarise(time_since_invasion = mean(time_since_invasion), .groups = "drop")

max_count <- max(ggplot_build(
  ggplot(lake_level_data, aes(x = time_since_invasion)) + 
    geom_histogram(binwidth = 1)
  )$data[[1]]$count, na.rm = TRUE)

all_species %>% 
  mutate(impact_groups = paste0(zm_lake, impact_yn)) %>% 
  group_by(dow, spec, impact_groups) %>% 
  summarise(time_since_invasion = mean(time_since_invasion)) %>% 
  ggplot(aes(x = time_since_invasion, fill = impact_groups)) +
  geom_histogram(binwidth = 1, alpha = 0.4, position = "identity", show.legend = FALSE) +
  geom_density(aes(y = after_stat(density) * max_count * 1, color = impact_groups), # Scaling density
               alpha = 0.01, linewidth = 1, show.legend = FALSE) +
  scale_y_continuous(
    name = "Histogram Count",
    sec.axis = sec_axis(~ . / max_count, name = "Density (scaled)")
  ) +
  geom_vline(xintercept = 0) +
  scale_color_manual("", values = c("lightblue", "lightblue", "salmon", "salmon"), labels = c("Uninvaded", "Uninvaded", "Zebra Mussel", "Zebra Mussel")) +
  scale_fill_manual("", values = c("lightblue", "lightblue", "salmon", "salmon"), labels = c("Uninvaded", "Uninvaded", "Zebra Mussel", "Zebra Mussel")) +
  theme_minimal() +
  facet_wrap(~spec, scales = "free")

rm(lake_level_data, all_species, max_count)
```

